---
title: "Task1"
author: "Andrii"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

### 1.1. Atomic Types

Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
#_code for solving task

#numeric
temperature <- 14
year <- 1223
elevation <- 118 #meters
distance <- 60 #kilometers


#integer
population <- 1620
area <- 17.5 #square kilometers
buildings <- 204


#character
city_name <- "Cierny Brod"
city_old_name <- "Kusoud"
district <- "Galanta district"


#logical
has_water <- TRUE
is_inhabited <- TRUE 
```



### 1.2. Vectors

Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(14, 100, by = 8)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 1.3. MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

```{r}
#_code for solving task
set.seed(123)  # so results stay the same each time
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)

# Name rows and columns according to your theme
rownames(my_matrix) <- c("Planet_A", "Planet_B", "Planet_C")
colnames(my_matrix) <- c("Energy", "Gravity", "Light")

# Display the matrix
my_matrix

# Calculate row and column averages
rowMeans(my_matrix)
colMeans(my_matrix)

```
### 1.4. LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
my_list <- list(
  # include some of your atomic variables
  population = population,
  area = area,
  has_water = has_water,
  city_name = city_name,
  
  # include your created vectors
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  
  # include your matrix
  matrix = my_matrix,
  
  # add a short description
  description = "This is my data universe — full of cities, planets, logic, and random numbers!"
)

# Display the list
my_list
```



### 1.5. Factor
Create a **factor variable** representing categories within your universe, e.g.:



- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task
energy_level <- factor(c("low", "medium", "high", "medium", "high", "low", "high"))

# Show counts of each level
table(energy_level)
#energy_level
```
### 1.6. Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   
```{r}
#_code for solving task
my_df <- data.frame(
  Planet = c("Astra", "Luna", "Zyra", "Kora", "Mira", "Vega"),
  Population = c(1620, 980, 1450, 720, 1230, 880),
  Habitable = c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE),
  Level = factor(c("high", "medium", "low", "medium", "high", "low"))
)

# Rename columns (if needed)
colnames(my_df) <- c("Planet_Name", "Population_Size", "Is_Habitable", "Development_Level")

# Display structure and summary
str(my_df)
summary(my_df)

# Add a new derived column: population ratio (relative to max)
my_df$Pop_Ratio <- my_df$Population_Size / max(my_df$Population_Size)

# Show final data frame
my_df
```

---
title: "Task2"
author: "Andrii"
date: "28/10/2025"
format:
  html:
    embed-resources: true
editor: source
---

## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors
Let's start with the vectors you created in **Task 1**
### 2.1.1. From your Sequence vector

# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 20
sequence_vector[sequence_vector > 20]

```{r}
# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 20
sequence_vector[sequence_vector > 20]
```


### 2.1.2. From your Logical vector

```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```

###  2.1.2. From your **Logical vector**:

- Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
- Use `which()` to get the indices (positions) of all the `TRUE` values.

```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```

### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

- Access and print the **description** string using the `$` operator
- Access and print your **3x3 matrix** using double brackets `[[ ]]`.
- Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
### Access and print specific elements from your list

# Access and print the description string
my_list$description

# Access and print your 3x3 matrix
my_list[[8]]   # or my_list$matrix

# Access and print the 2nd element from your numeric vector within the list
my_list$numeric_vector[2]
```
### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

- Print the **first 3 rows**. (Use `head()`.)
- Print the value from the **4th row, 1st column**.
- Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

- Example: `df_subset <- your_df[your_df$population > 1000, ]`

`%in%` **Subsetting**: Create a new data frame `df_subset_2` that selects rows belonging to **two specific categories** from your factor or character column.

- Example: `df_subset_2 <- your_df[your_df$region %in% c("North", "South"), ]`

```{r}
### Print the first 3 rows
head(my_df, 3)

### Print the value from 4th row, 1st column
my_df[4, 1]

### Print only the logical column (Is_Habitable)
my_df$Is_Habitable


### Logical Subsetting: keep only planets with population > 1000
df_subset <- my_df[my_df$Population_Size > 1000, ]
df_subset


### %in% Subsetting: keep planets with "high" or "medium" development
df_subset_2 <- my_df[my_df$Development_Level %in% c("high", "medium"), ]
df_subset_2
```
### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

- Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

- Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
- Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

- Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

- Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.

```{r}
### Create a copy
df_corrupted <- my_df

### Introduce 3 NAs
df_corrupted[2, 2] <- NA
df_corrupted[5, 3] <- NA
df_corrupted[6, 4] <- NA

### Find total number of missing values
sum(is.na(df_corrupted))

### Find how many NAs per column
colSums(is.na(df_corrupted))

### Calculate mean of Population_Size with NAs (will return NA)
mean(df_corrupted$Population_Size)

### Calculate mean while ignoring NAs
mean(df_corrupted$Population_Size, na.rm = TRUE)

### Remove rows with NAs
df_clean <- na.omit(df_corrupted)
df_clean
```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

Use that index to print the row with that observation.

- Example: `highest_pop_index <- which.max(your_df$population)`
- Example: `your_df[highest_pop_index, ]`
```{r}
### Find row with highest population
highest_pop_index <- which.max(my_df$Population_Size)
my_df[highest_pop_index, ]

### Find row with lowest population
lowest_pop_index <- which.min(my_df$Population_Size)
my_df[lowest_pop_index, ]
```

