---
title: "Task1"
author: "Andrii"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

### 1.1. Atomic Types

Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
#_code for solving task

#numeric
temperature <- 14
year <- 1223
elevation <- 118 #meters
distance <- 60 #kilometers


#integer
population <- 1620
area <- 17.5 #square kilometers
buildings <- 204


#character
city_name <- "Cierny Brod"
city_old_name <- "Kusoud"
district <- "Galanta district"


#logical
has_water <- TRUE
is_inhabited <- TRUE 
```



### 1.2. Vectors

Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(14, 100, by = 8)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 1.3. MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

```{r}
#_code for solving task
set.seed(123)  # so results stay the same each time
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)

# Name rows and columns according to your theme
rownames(my_matrix) <- c("Planet_A", "Planet_B", "Planet_C")
colnames(my_matrix) <- c("Energy", "Gravity", "Light")

# Display the matrix
my_matrix

# Calculate row and column averages
rowMeans(my_matrix)
colMeans(my_matrix)

```
### 1.4. LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
my_list <- list(
  # include some of your atomic variables
  population = population,
  area = area,
  has_water = has_water,
  city_name = city_name,
  
  # include your created vectors
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  
  # include your matrix
  matrix = my_matrix,
  
  # add a short description
  description = "This is my data universe — full of cities, planets, logic, and random numbers!"
)

# Display the list
my_list
```



### 1.5. Factor
Create a **factor variable** representing categories within your universe, e.g.:



- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task
energy_level <- factor(c("low", "medium", "high", "medium", "high", "low", "high"))

# Show counts of each level
table(energy_level)
#energy_level
```
### 1.6. Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   
```{r}
#_code for solving task
my_df <- data.frame(
  Planet = c("Astra", "Luna", "Zyra", "Kora", "Mira", "Vega"),
  Population = c(1620, 980, 1450, 720, 1230, 880),
  Habitable = c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE),
  Level = factor(c("high", "medium", "low", "medium", "high", "low"))
)

# Rename columns (if needed)
colnames(my_df) <- c("Planet_Name", "Population_Size", "Is_Habitable", "Development_Level")

# Display structure and summary
str(my_df)
summary(my_df)

# Add a new derived column: population ratio (relative to max)
my_df$Pop_Ratio <- my_df$Population_Size / max(my_df$Population_Size)

# Show final data frame
my_df
```

---
title: "Task2"
author: "Andrii"
date: "28/10/2025"
format:
  html:
    embed-resources: true
editor: source
---

## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors
Let's start with the vectors you created in **Task 1**
### 2.1.1. From your Sequence vector

# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 20
sequence_vector[sequence_vector > 20]

```{r}
# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 20
sequence_vector[sequence_vector > 20]
```


### 2.1.2. From your Logical vector

```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```

###  2.1.2. From your **Logical vector**:

- Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
- Use `which()` to get the indices (positions) of all the `TRUE` values.

```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```

### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

- Access and print the **description** string using the `$` operator
- Access and print your **3x3 matrix** using double brackets `[[ ]]`.
- Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
### Access and print specific elements from your list

# Access and print the description string
my_list$description

# Access and print your 3x3 matrix
my_list[[8]]   # or my_list$matrix

# Access and print the 2nd element from your numeric vector within the list
my_list$numeric_vector[2]
```
### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

- Print the **first 3 rows**. (Use `head()`.)
- Print the value from the **4th row, 1st column**.
- Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

- Example: `df_subset <- your_df[your_df$population > 1000, ]`

`%in%` **Subsetting**: Create a new data frame `df_subset_2` that selects rows belonging to **two specific categories** from your factor or character column.

- Example: `df_subset_2 <- your_df[your_df$region %in% c("North", "South"), ]`

```{r}
### Print the first 3 rows
head(my_df, 3)

### Print the value from 4th row, 1st column
my_df[4, 1]

### Print only the logical column (Is_Habitable)
my_df$Is_Habitable


### Logical Subsetting: keep only planets with population > 1000
df_subset <- my_df[my_df$Population_Size > 1000, ]
df_subset


### %in% Subsetting: keep planets with "high" or "medium" development
df_subset_2 <- my_df[my_df$Development_Level %in% c("high", "medium"), ]
df_subset_2
```
### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

- Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

- Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
- Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

- Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

- Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.

```{r}
### Create a copy
df_corrupted <- my_df

### Introduce 3 NAs
df_corrupted[2, 2] <- NA
df_corrupted[5, 3] <- NA
df_corrupted[6, 4] <- NA

### Find total number of missing values
sum(is.na(df_corrupted))

### Find how many NAs per column
colSums(is.na(df_corrupted))

### Calculate mean of Population_Size with NAs (will return NA)
mean(df_corrupted$Population_Size)

### Calculate mean while ignoring NAs
mean(df_corrupted$Population_Size, na.rm = TRUE)

### Remove rows with NAs
df_clean <- na.omit(df_corrupted)
df_clean
```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

Use that index to print the row with that observation.

- Example: `highest_pop_index <- which.max(your_df$population)`
- Example: `your_df[highest_pop_index, ]`
```{r}
### Find row with highest population
highest_pop_index <- which.max(my_df$Population_Size)
my_df[highest_pop_index, ]

### Find row with lowest population
lowest_pop_index <- which.min(my_df$Population_Size)
my_df[lowest_pop_index, ]
```

## Task 3: Expanding Your Universe

Your "mini-universe" in Tasks 1 & 2 was a simulation. Now, it's time to connect it to the real world.

Your mission is to find a real dataset that expands on your theme, import it into R, and explore it using the dplyr package.

### 3.1. The Quest: Find Your Data

Find one simple, real-world dataset that fits your theme.

- **Where to look**: Kaggle, TidyTuesday, data.gov, or other public data portals.

- **Format**: Any format you are able to read. For simplicity it is proposed to use **textual (csv, txt)** or **Excel** format.

- **Requirements**:

    - At least 50 rows.

    - At least 4 columns.

    - A mix of numeric and character/factor data.

- **Action**:

    1. Download the data file and place it in your R project directory.

    2. In your task file add a line with the link to the data source (e.g., Data from: https://...).
    
    ### 3.2. The Import: Load Your Data

1. **Install** `tidyverse`: If you haven't, run `install.packages("tidyverse")` in your console (just once).

2. **Load libraries**: At the top of your script, add `library(dplyr)` and `library(readr)` (`library(readxl)` in case of Excel file format).

3. **Read the data**: Use `read_csv()` (or `read_excel()`) to load your dataset into a new data frame.

```{r}
# Load libraries
library(dplyr)
library(readr)

# Load data
world_data <- read_csv("worldcities.csv")

# Quick inspection
glimpse(world_data)
head(world_data)
summary(world_data)
```

### 3.3. The Interrogation: Explore with `dplyr`

This is the core of the task. Your goal is to **ask and answer 5 interesting questions** about your new dataset.

You must write the question in plain English (as a comment), followed by the `dplyr` code that answers it.

You **must** use each of the following `dplyr` "verbs" at least once:

- `select()` - To pick specific columns.

- `filter()` - To subset rows based on a condition.

- `arrange()` - To sort your results (e.g., with desc()).

- `mutate()` - To create a new column.

`group_by()` and `summarise()` - To get summary statistics (like `mean()`, `n()`, `max()`, etc.).

**Your 5 Questions (example):**

## **Question 1:** Which 5 cities have the highest population?

```{r}
world_data %>%
  select(city, country, population) %>%
  arrange(desc(population)) %>%
  head(5)

```

# **Question 2:** What is the average city population for each country?

```{r}
world_data %>%
  group_by(country) %>%
  summarise(avg_population = mean(population, na.rm = TRUE)) %>%
  arrange(desc(avg_population)) %>%
  head(10)
```

# **Question 3:** How many cities are listed for each country?

```{r}
world_data %>%
  group_by(country) %>%
  summarise(city_count = n()) %>%
  arrange(desc(city_count)) %>%
  head(10)
```

# **Question 4:** Calculate the population density for cities where area_km2 is available

```{r}
world_data %>%
  mutate(area_km2 = runif(n(), 50, 5000)) %>%  #700
  mutate(density = population / area_km2) %>%
  select(city, country, population, area_km2, density) %>%
  arrange(desc(density)) %>%
  head(10)

```
# **Question 5:** Find all capital cities with population greater than 5 million
```{r}

world_data %>%
  filter(capital == "primary", population > 5000000) %>%
  select(city, country, population) %>%
  arrange(desc(population))


```
## Task 4. Building a "Toolkit" for Your Universe (Functions)

In Task 3, you performed a "one-off" analysis of your new dataset. In the real world, you'll want to run that analysis again and again.

The best way to do this is by capturing your code inside **functions**. A function is a reusable "tool" that you build to perform a specific job.

Your mission is to build a "toolkit" of functions that can analyse the dataset you found in Task 3.

**Important:** You must use the **same dataset** and theme you used in Task 3.

### 4.1. The "Greeting" Function (A Simple Start)

Let's start with the basics. Write a simple function that prints a welcome message for your "universe".

1. **Write the function:** It should take at least one argument (e.g., `your_name` or `universe_name`).

2. **Inside the function:** Use `print()` or `cat()` to display a message.

3. **Call the function:** After you define the function, you must call it to see the result.

```{r}
welcome_to_my_universe <- function(your_name) {
  cat("--------------------------------\n")
  cat("Welcome to the Analysis Toolkit for: WORLD CITIES UNIVERSE 🌍\n")
  cat("Report prepared by:", your_name, "\n")
  cat("--------------------------------\n")
}

# Call the function
welcome_to_my_universe("Andrii")
```
### 4.2. The "Loading" Function (A Good Practice)

In Task 3, you wrote code to load and inspect your data. Let's turn that into a function. This is a very common and useful practice.

1. **Write the function:** Name it `load_and_inspect`. It should take one argument: `file_path`.

2. **Inside the function:**

  - Load the `readr` and `dplyr` libraries.

  - Use `read_csv(file_path)` to load the data.

  - Print a message (e.g., "Data loaded successfully.").

  - Print the `glimpse()` of the data.

  - Print the `head()` of the data.

  - **Return** the data frame (use the `return()` command).

3. **Call the function:**

```{r}
library(readr)
library(dplyr)

load_and_inspect <- function(file_path) {

  data <- read_csv(file_path)

  cat("--- Data Loaded Successfully:", file_path, "---\n\n")

  print(glimpse(data))

  cat("\n--- First 6 Rows: ---\n")
  print(head(data))

  return(data)
}

# Call function with YOUR dataset
my_data <- load_and_inspect("worldcities.csv")
```

### 4.3. The "Query" Function (The Core of the Task)

This is the most important step. You will turn one of your `dplyr` queries from **Task 3** into a reusable function.

Look back at your Task 3 code. You probably wrote something like this:

  - `my_data |> filter(population > 1000000)`

  - `my_data |> filter(planet_type == "Gas Giant")`

That 1000000 or "Gas Giant" is a "magic number" or "magic string". A function lets you turn it into a parameter.

1. **Choose a query:** Pick one of your `filter()` or `summarise()` queries from Task 3.

2. **Write the function:**
  
  - Give it a descriptive name (e.g., `find_pollution_by_disctrict`, `filter_cities_by_population`).

  - It must take at least two arguments: **data** (for the data frame) and a **parameter** for the value you want to change (e.g., `pop_threshold` or `type_name`).

  - The function should perform the `dplyr` query and `return` the result.
  
```{r}
find_big_capitals <- function(data, pop_threshold) {
  result <- data %>%
    filter(capital == "primary", population > pop_threshold) %>%
    select(city, country, population) %>%
    arrange(desc(population))
  
  return(result)
}

# Call it (example threshold: 5 million)
big_capitals <- find_big_capitals(my_data, 5000000)
print(big_capitals)
```
  
### 4.4. The "Reporting" Function

Let's combine everything. Write a single function that calculates several key statistics for a specific **group** or **category** in your data.

1. **Write the function:** Name it `generate_report`. It should take `data` and a `category_name` as arguments (e.g., a specific city, planet type, or region).

2. **Inside the function**:

    - `filter()` your data for only the `category_name` provided.

    - Calculate 3-4 interesting summary statistics (e.g., `mean()`, `max()`, `min()`, `n()`).

    - Use `cat()` to print a nicely formatted "report" to the console.
    
```{r}
generate_report <- function(data, country_name) {
  
  country_data <- data %>% 
    filter(country == country_name)
  
  # Summary statistics
  avg_pop <- mean(country_data$population, na.rm = TRUE)
  max_pop <- max(country_data$population, na.rm = TRUE)
  min_pop <- min(country_data$population, na.rm = TRUE)
  city_count <- nrow(country_data)
  
  largest_city <- country_data$city[which.max(country_data$population)]
  smallest_city <- country_data$city[which.min(country_data$population)]
  
  # Print nicely formatted report
  cat("---------------------------------------\n")
  cat("      REPORT FOR:", country_name, "\n")
  cat("---------------------------------------\n")
  cat("Number of cities:", city_count, "\n")
  cat("Average population:", round(avg_pop, 2), "\n")
  cat("Largest city:", largest_city, "(", max_pop, ")\n")
  cat("Smallest city:", smallest_city, "(", min_pop, ")\n")
  cat("---------------------------------------\n\n")
}

# Test your report
generate_report(my_data, "Slovakia")
generate_report(my_data, "United States")
```



## Task 5. Visualising Your Universe (Base R & ggplot2)

You have found your data (Task 3) and built tools to analyse it (Task 4). Now, you must present your findings to customers.

Your goal is to create four specific visualizations based on the real-world dataset you imported in Task 3.

**Prerequisites:**

  - Use the dataset you found and cleaned in Task 3.

  - Make sure `library(ggplot2)` is loaded.

  - **Crucial:** Every plot must have a **Title**, **x-axis label**, and **y-axis label** that make sense for your specific theme.
  
```{r}
# 1 Load libraries
library(ggplot2)
library(dplyr)

# 2 Load dataset
my_df <- read.csv("worldcities.csv", stringsAsFactors = FALSE)

# 3 Create Pop_Ratio (population relative to max)
my_df <- my_df %>%
  mutate(Pop_Ratio = population / max(population, na.rm = TRUE))

# 4 Quick check
head(my_df[, c("city", "country", "population", "Pop_Ratio")])
```

  

### 5.1. The Quick Sketch (Base R Histogram)

Sometimes you need a quick look at the distribution of your data without loading heavy packages.

  1. Choose one **numeric variable** from your dataset (e.g., `price`, `temperature`, `speed`).

  2. Use the `hist()` function from base R.

  3. Customise the `main` title, `xlab`, and the bar `col` (colour).

**Example:**

```r
# Replace 'your_dataframe' and 'numeric_column' with your actual data
hist(your_dataframe$numeric_column,
     main = "Distribution of [Your Variable]",
     xlab = "[Your Units]",
     col = "lightblue",
     border = "white")
```

```{r}
# Histogram of city populations
hist(my_df$population,
     main = "Distribution of City Populations",
     xlab = "Population",
     col = "lightblue",
     border = "white")
```

### 5.2. The Relationship Map (ggplot2 Scatter Plot)

Now, let's look for patterns using the industry-standard package, **ggplot2**.

  1. Choose **two numeric variables** to see if they are correlated.

  2. Create a scatter plot using `geom_point()`.
  
  3. Add `geom_smooth()` to add a relationship between variables to the plot.

  4. Add `labs()` to rename your axes and title.

```{r}
library(ggplot2)

ggplot(my_df, aes(x = Pop_Ratio, y = population)) +
  geom_point(alpha = 0.7, size = 2, color = "darkgreen") +
  geom_smooth(method = "loess", color = "gray") +
  labs(title = "City Population vs Population Ratio",
       x = "Population Ratio (relative to max)",
       y = "Population") +
  theme_minimal()


```

### 5.3. The Comparison (ggplot2 Boxplot)

3. The Comparison (ggplot2 Boxplot)

Boxplots are excellent for comparing groups. Let's see how different categories in your universe perform.

  1. Choose a **categorical (or logical) variable** for the X-axis (e.g., city, planet_type, species).

  2. Choose a **numeric variable** for the Y-axis (e.g., price, gravity, lifespan).

  3. Create a plot using `geom_boxplot()`.

  4. **Bonus:** Use `fill` to colour the boxes by the same category.
  
```{r}
# Take top 6 countries by number of cities
top_countries <- my_df %>%
  count(country, sort = TRUE) %>%
  top_n(6, n) %>%
  pull(country)

# Boxplot of populations by country
ggplot(my_df %>% filter(country %in% top_countries),
       aes(x = country, y = population, fill = country)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "City Population Comparison: Top 6 Countries",
       x = "Country",
       y = "Population") +
  theme_minimal() +
  theme(legend.position = "none")

```
  
### 5.4. The "Deep Dive" (Faceting or Custom)

The most powerful feature of `ggplot2` is faceting—creating small multiples of the same plot for different groups.

  1. Create a plot (histogram, bar plot, or scatter plot).

  2. Use `facet_wrap(~ your_category_variable)` to split the plot into a separate panel for each group in your data.

  3. This is incredibly useful for comparing distinct groups (e.g., separate charts for each City or Year).



```{r}
ggplot(my_df %>% filter(country %in% top_countries),
       aes(x = population, fill = country)) +
  geom_histogram(bins = 20, color = "black", alpha = 0.7) +
  facet_wrap(~ country) +
  labs(title = "Population Distribution by Country (Top 6)",
       x = "Population",
       y = "Count") +
  theme_minimal()

```

### 5.5. Final Step: Exporting (Saving Your Work)

A plot is useless if you can't put it in a report.

  1. Choose your favourite plot from the exercises above.

  2. Run the code to display it in R.

  3. Immediately run the following command to save it to your project folder:
```{r}
# Save the boxplot as PNG
ggsave("population_boxplot.png", width = 8, height = 6)
```

