---
title: "Task1"
author: "Andrii"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

### 1.1. Atomic Types

Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
#_code for solving task

#numeric
temperature <- 14
year <- 1223
elevation <- 118 #meters
distance <- 60 #kilometers


#integer
population <- 1620
area <- 17.5 #square kilometers
buildings <- 204


#character
city_name <- "Cierny Brod"
city_old_name <- "Kusoud"
district <- "Galanta district"


#logical
has_water <- TRUE
is_inhabited <- TRUE 
```



### 1.2. Vectors

Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(14, 100, by = 8)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 1.3. MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

```{r}
#_code for solving task
set.seed(123)  # so results stay the same each time
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)

# Name rows and columns according to your theme
rownames(my_matrix) <- c("Planet_A", "Planet_B", "Planet_C")
colnames(my_matrix) <- c("Energy", "Gravity", "Light")

# Display the matrix
my_matrix

# Calculate row and column averages
rowMeans(my_matrix)
colMeans(my_matrix)

```
### 1.4. LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
my_list <- list(
  # include some of your atomic variables
  population = population,
  area = area,
  has_water = has_water,
  city_name = city_name,
  
  # include your created vectors
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  
  # include your matrix
  matrix = my_matrix,
  
  # add a short description
  description = "This is my data universe — full of cities, planets, logic, and random numbers!"
)

# Display the list
my_list
```



### 1.5. Factor
Create a **factor variable** representing categories within your universe, e.g.:



- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task
energy_level <- factor(c("low", "medium", "high", "medium", "high", "low", "high"))

# Show counts of each level
table(energy_level)
#energy_level
```
### 1.6. Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   
```{r}
#_code for solving task
my_df <- data.frame(
  Planet = c("Astra", "Luna", "Zyra", "Kora", "Mira", "Vega"),
  Population = c(1620, 980, 1450, 720, 1230, 880),
  Habitable = c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE),
  Level = factor(c("high", "medium", "low", "medium", "high", "low"))
)

# Rename columns (if needed)
colnames(my_df) <- c("Planet_Name", "Population_Size", "Is_Habitable", "Development_Level")

# Display structure and summary
str(my_df)
summary(my_df)

# Add a new derived column: population ratio (relative to max)
my_df$Pop_Ratio <- my_df$Population_Size / max(my_df$Population_Size)

# Show final data frame
my_df
```


